A Mechanised Library for Generic (Co)Programming

### 2024-02-28 paper/paper.tex
Stuff
Abreu et al POP2023 extraction

If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$
does have a unique solution.  We show this by first providing the definition
for recursive hylomorphisms in Coq, and then proving that any function
satisfying this equation is a hylomorphism. We call these hylomorphisms with a
recursive coalgebra, \textbf{recursive hylomorphisms}.
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H 
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \coq{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{coqcode}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for 
hylomorphism fusion as an example:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
  (g2 : Alg F C) (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
  : f2 \o hylo g1 h1 =e hylo g2 h1.
Proof.
  rewrite hylo_univ, fmap_comp compA <- E2 <- compA <- compA (compA g1) <- hylo_unr
  reflexivity.
Qed.
  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
\end{coqcode}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}, where two consecutive recursive computations, one that builds a
data structure, and another one that consumes it, can be fused together into a
single recursive definition. This, in turn, allows us to prove that a recursive
hylomorphism is the composition of a catamorphism and a recursive anamorphism.
\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

@@

\dcas{Talk About this}
\begin{minted}{OCaml}
  let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{minted}

>>>>

If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$
does have a unique solution.  We show this by first providing the definition
for recursive hylomorphisms in Coq, and then proving that any function
satisfying this equation is a hylomorphism. We call these hylomorphisms with a
recursive coalgebra, \textbf{recursive hylomorphisms}.
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H 
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \coq{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{coqcode}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for 
hylomorphism fusion as an example:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) 
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.

  rewrite hylo_univ, fmap_comp, compA, <- E2, 
          <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
Qed.
  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
\end{coqcode}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}, where two consecutive recursive computations, one that builds a
data structure, and another one that consumes it, can be fused together into a
single recursive definition. This, in turn, allows us to prove that a recursive
hylomorphism is the composition of a catamorphism and a recursive anamorphism.
\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

@@

This section covers how to use our approach to calculate recursive programs
from simple, non-recursive specifications. The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.

\subsection{Sorting Algorithms}
\dcas{Talk About this}
\begin{minted}{OCaml}
  let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{minted}

\subsection{Knapsack}

\subsection{Harper's Regular-Expression Matcher}
